Methodology for Assessing the State of the Practice for Domain X

### 2020-12-18 StateOfPractice/Methodology/Methodology.tex
Completed revision of introduction of methodology section

Research software uses computing to simulate mathematical models of real world
systems so that we can better understand and predict those system's behaviour.  A
small set of examples of important research software includes the
following: designing new automotive parts, analyzing the flow of blood in the
body, and determining the concentration of a pollutant released into the
groundwater.  As these examples illustrate, research software can be used for
tackling important problems that impact such areas as manufacturing, financial
planning, environmental policy, and the health, welfare and safety of
communities.

@@

The purpose of our proposed methodology is to gain insight on how software
quality is impacted by the software development principles, processes and tools
currently used within research software communities.  Since research software is
so broad a category, we will reduce the scope of each application of the
methodology by applying it to a specific domain within research software.  This
``state of the practice for domain X'' exercise will build off of prior work on
measuring/assessing the state of software development practice in several
research domains.  We will update the work that was done previously for domains
such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh
Generators \citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.

In the previous project, we measured 30 software projects for each domain, but
the measures were relatively shallow.  With this re-boot we are adding more
measures and experiments.  refining the list
into a short list of the best projects for more detailed measurement.  More details.

In the previous state of the practice assessing exercise
\citep{SmithEtAl2018_arXivGIS}, a series of questions and simple metrics was
devised to measure the quality of the documentation and adherence to best
software development practices.  The previous set of questions is provided
\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
{here}. As described below, the new project critically assess the previous set of questions and
revises as necessary.  In addition, the following data is also collected for each domain:

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain.  What services does each member of the domain provide?  This
  information would be summarized in a commonality analysis document.
\item Usability testing of each software in the domain.
\item Collect empirical software engineering related data, such as the number of
  files, number of lines of code, number of open issues,
  number of closed issues, the number of lines of documentation (?). 
\end{itemize}

One of the guidelines for successfully on-boarding new developers is to
  ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
  selected measurements should be able to assist with judging whether a project
  succeeds on this front.

The above data will be combined to rank the software in the domain.  The
specifics are yet to be determined, but in the past the Analytic Hierarchy
Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
results were calculated by a grade out of 10 from each student evaluator for
each quality for each measured product.  A better approach may be to have a
group of people familiar with the software do a pairwise comparison between each
product to come up with a number.  The number would be based on the data
collected, but not in a numerical way.  The data collected would serve the same
role as a decision support system.  It would provided insight, but human beings
would do the final pairwise comparisons, and thus the final ranking.

Key to the success of this exercise will be to involve and engage a partner for
each state of practice project.  In the previous effort we did not involve domain
experts with the rationale of excluding potential bias, but this advantage is
not worth not being able to evaluate the functionality/usability of the software.
Moreover, not having an expert makes publication more difficult, since there is
no one to advice on how best to approach journals and publishers.  The domain
expert will be asked to help in the following ways:

GitHub will be used to coordinate the work of the large team of people that will
be involved in this project.  In addition to the project record left on GitHub,
the final data will be exported to Mendeley (as was done for the previous
quality measurement exercise, for instance the grades for GIS software are
summarized at \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
{https://data.mendeley.com/datasets/6kprpvv7r7/1}.)

The data for each of the different subdomains within SCS will be standardized to
facilitate a meta-analysis that combines all of the data.

The scope of this document is to conduct a current state of practice analysis of
software in a scientific computing domain for the purpose of understanding
quality and assisting in future development. We observe product, artifact, and
process quality in our analysis. We leave the assessment of scientific computing
software using performance benchmarks to other projects, such as the work of
\cite{kaagstrom1998gemm}. The methodology presented here is part of an overall
assessment of the impact of model driven engineering and code generation on the
sustainability of scientific computing software. The research proposal for this
assessment, which contains a more detailed description of this state of practice
exercise, can be found
\href{https://github.com/smiths/AIMSS/blob/master/OverallResearchProposal/ResearchProposal.pdf}
{here}. The research questions that this exercise will answer are found in
Section \ref{ResearchQuestions}. Section \ref{StepsAQDS} outlines the steps to
be taken.

@@

\begin{enumerate}
\item The domain must have well-defined and stable theoretical underpinning.  A
  good sign of this is the existance of standard textbooks, preferably
  understandable by an upper year undergraduate student.
\item There must be a community of people studying the domain.
\item The software packages must have open source options.
\item A preliminary search, or discussion with experts, suggests that there will
  be numerous, at least 15, candidate software packages in the domain that
  qualify as `research software'.
\end{enumerate}

>>>>

Research software uses computing to simulate mathematical models of real world
systems so that we can better understand and predict those systems's behaviour.  A
small set of examples of important research software includes the
following: designing new automotive parts, analyzing the flow of blood in the
body, and determining the concentration of a pollutant released into the
groundwater.  As these examples illustrate, research software can be used for
tackling important problems that impact such areas as manufacturing, financial
planning, environmental policy, and the health, welfare and safety of
communities.

@@

The purpose of our proposed methodology is to understand how software quality is
impacted by the software development principles, processes and tools currently
used within research software communities.  Since research software is so broad
a category, we will reduce the scope of our methodology to focus on one specific
research software domain at a time.  This ``state of the practice for domain X''
exercise will build off of prior work on measuring/assessing the state of
software development practice in several research domains.  We will update the
work that was done previously for domains such as Geographic Information Systems
\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Seismology software \citep{SmithEtAl2018}, and Statistical software for
psychology \citep{SmithEtAl2018_StatSoft}.

In the previous ``state of the practice'' project, we measured 30 software
projects for each domain, but the measures were relatively shallow.  With this
re-boot we will still target about 30 software examples from each domain, but we
will be collecting more data by adding more measures and experiments.  In
keeping with the previous project, we still have the constraint that the work
load for applying the methodology to a given domain needs to be feasible for a
team as small as one individual, and for a time that is short, ideally 3 to 4
months of part-time work per domain.

To start with, the 
\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
{previous set of questions} were critically assessed and modified.  In addition,
the following data is part of the new methodology:

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain via a commonality analysis.
\item Usability testing for a sample of software in the domain.
\item Empirical software engineering related data, such as the number of
  files, number of lines of code, percentage of issues that are closed, etc. 
\end{itemize}

Unlike the previous measurement process, the new methodology will involves and
engages a domain expert partner throughout.  We did not previously engage the
domain expert with the rationale that we wished to exclude potential bias.
However, this advantage is not worth not being able to evaluate the
functionality/usability of the software.  Moreover, not having an expert makes
publication more difficult, since there is no one to provide advice on how best
to approach journals and publishers.

In the proposed methodology, the collected data will be combined to rank the
software within each domain using the Analytic Hierarchy Process (AHP)
\citep{Saaty1980}.  As in the past process, we will use AHP to develop a list of software
ranked by quality.  However, in the new process we will not stop with this
list.  The domain expert will be consulted to verify the ordering, and to
discuss the decisions that led to the ranking.  The AHP process will be used to
facilitate a conversation with the domain experts as a means to deepen our
understanding of the software in the domain, and the needs of typical developers.

GitHub will be used to coordinate the work of the team of people that will
be involved in this project.  In addition to the project record left on GitHub,
the final data will be exported to Mendeley (as was done for the previous
quality measurement exercise, for instance the grades for GIS software are
summarized at \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
{https://data.mendeley.com/datasets/6kprpvv7r7/1}.)

The scope of this methodology includes observations on product, artifact and
process quality for research softwares. We leave the assessment of research
software using performance benchmarks to other projects, such as the work of
\cite{kaagstrom1998gemm}.

@@

\begin{enumerate}
\item The domain must have well-defined and stable theoretical underpinning.  A
  good sign of this is the existence of standard textbooks, preferably
  understandable by an upper year undergraduate student.
\item There must be a community of people studying the domain.
\item The software packages must have open source options.
\item A preliminary search, or discussion with experts, suggests that there will
  be numerous, at least 15, candidate software packages in the domain that
  qualify as `research software'.
\end{enumerate}
